---
title: "Gene names in different databases"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
fig_caption: yes
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Gene names in different databases}
  %\usepackage[UTF-8]{inputenc}
---

```{r, echo = FALSE, message = FALSE}
library(phylomapr)
library(dplyr)
library(biomartr)
options(width = 750)
knitr::opts_chunk$set(
  comment = "#>",
  error = FALSE,
  tidy = FALSE)
```

## Motivation

The GeneID labels differ across databases. This is a potential problem when gene age inference has been performed from protein sequences obtained from one database (e.g. UniProtKB) and the RNA-seq quantification has been performed on genes from another database (e.g. ENSEMBL).

## Solution

In `phylomapr`, we try to mitigate this issue using the helper function `convertID()`, which essentially wraps around the function `biomart()` in the [`biomartr`](https://docs.ropensci.org/biomartr/) package. In this article, we give an example of how _C. elegans_ phylomaps can be converted from the uniprot format to the wormbase format.

Typically, the gene IDs can be converted to the ENSEMBLE gene IDs, whose `biomartr` attributes are `ensembl_gene_id` and `ensembl_peptide_id` by default.

```{r, message = FALSE, results='hide', warning = FALSE, eval=FALSE}
# default run of convertID
Caenorhabditis_elegans.PhyloMap.ENSEMBL <- 
  phylomapr::convertID(
    phylomap = phylomapr::Caenorhabditis_elegans.PhyloMap,
    mart = "ENSEMBL_MART_ENSEMBL",
    dataset = "celegans_gene_ensembl",
    filters = "uniprot_gn_id"
    )
```

If you run this, the gene IDs are named as e.g. `WBGene00000001`. However, gene IDs in other RNA-seq experiments employ other gene names. In the case of _C. elegans_ this may be wormbase IDs (e.g. `2L52.1a`). How can we convert the phylomaps to accommodate this?

First, the `organismFilters` function in `biomartr` can be used to obtain the attributes. Here, we ask for the topic `worm` because we are interested in wormbase IDs

```{r, message = FALSE, results='hide', warning = FALSE}
c_elegans_BM.filt <- organismFilters(organism = "Caenorhabditis elegans", topic = "worm")
```
```{r}
c_elegans_BM.filt
```

After scrolling through the results, we observe that the attribute `wormbase_cds` should work as the IDs are formatted in the `2L52.1a.1`-form.

Next, we convert the GeneID labels from uniprot to wormbase using `"wormbase_cds"` as the attribute.

```{r}
# here we use the dataset phylomapr::Caenorhabditis_elegans.PhyloMap
Celegans.PhyloMap <- phylomapr::Caenorhabditis_elegans.PhyloMap

Celegans.PhyloMap.wormbase <-
  phylomapr::convertID(
    phylomap = Celegans.PhyloMap,
    mart = "ENSEMBL_MART_ENSEMBL",
    dataset = "celegans_gene_ensembl",
    attributes = c("wormbase_cds"),
    filters    = "uniprot_gn_id")
```

```r
> Celegans.PhyloMap
# A tibble: 19,818 × 2
   Phylostratum GeneID                   
          <dbl> <chr>                    
 1            2 sp|A0A061ACU2|PIEZ1_CAEEL
 2            3 sp|A0A078BQN7|FKH3_CAEEL 
 3            1 sp|A0A078BQP2|GCY25_CAEEL
 4           10 sp|A0A0K3AUE4|SEA2_CAEEL 
 5            1 sp|A0A0K3AUJ9|PRDX_CAEEL 
 6            1 sp|A0A0K3AV08|MLK1_CAEEL 
 7            1 sp|A0A0K3AWM6|MOM5_CAEEL 
 8            3 sp|A0A0K3AXH1|ARID1_CAEEL
 9            1 sp|A0A131MBU3|IRG7_CAEEL 
10            1 sp|A0A131MCZ8|CNNM3_CAEEL
# ℹ 19,808 more rows
# ℹ Use `print(n = ...)` to see more rows
```

```r
> Celegans.PhyloMap.wormbase
# A tibble: 16,319 × 2
   Phylostratum GeneID  
          <dbl> <chr>   
 1            2 2RSSE.1a
 2            2 2RSSE.1b
 3            2 2RSSE.1c
 4           14 3R5.1a  
 5           14 3R5.1b  
 6            1 4R79.1a 
 7            1 4R79.1b 
 8            1 AC3.10  
 9           11 AC3.1a  
10           11 AC3.1b  
# ℹ 16,309 more rows
# ℹ Use `print(n = ...)` to see more rows
```

Tada! While the conversion issue between databases are mitigated, there are still some aspects to be improved. For example, we can see a drop in the number of genes in the phylomap. Furthermore, a bit of fidgeting with `biomartr` is needed. However, this should still be faster than running gene age inference from scratch!
